---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mori

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/mori)](https://CRAN.R-project.org/package=mori)
[![R-CMD-check](https://github.com/shikokuchuo/mori/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/shikokuchuo/mori/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/shikokuchuo/mori/graph/badge.svg)](https://app.codecov.io/gh/shikokuchuo/mori)
<!-- badges: end -->

      ________
     /\ mori  \
    /  \       \
    \  /  森   /
     \/_______/

Shared Memory for R Objects


→ `share()` writes an R object into shared memory and returns a shared version

→ ALTREP serialization hooks — shared objects serialize compactly and work transparently with `serialize()` and `mirai()`

→ ALTREP-backed lazy access — a 100-column data frame is one `mmap`; columns materialize on first touch

→ OS-level shared memory (POSIX / Win32) — pure C, no external dependencies; read-only in other processes, preventing corruption of shared data

→ Automatic cleanup — shared memory is freed when the R object is garbage collected

<br />

## Why mori

Parallel computing multiplies memory.
When 8 workers each need the same 200 MB dataset, that is 1.6 GB of serialization, transfer, and deserialization — with 8 separate copies consuming RAM.

`share()` writes the data into shared memory once and each worker maps the same physical pages — turning per-worker copies into per-worker references.

```{r}
#| label: bench-setup
library(mori)
library(mirai)
library(lobstr)

daemons(8)

# 200 MB data frame — 5 columns × 5M rows
df <- as.data.frame(matrix(rnorm(25e6), ncol = 5))
shared_df <- share(df)
```

Without mori, each worker holds the full data frame.
With mori, each worker holds a small reference into the shared region:

```{r}
#| label: bench-mem
mirai_map(1:8, \(i, data) format(lobstr::obj_size(data)),
          .args = list(data = df))[.flat] |> unique()

mirai_map(1:8, \(i, data) format(lobstr::obj_size(data)),
          .args = list(data = shared_df))[.flat] |> unique()
```

Avoiding 8 × 200 MB of serialize / deserialize also translates into a significant runtime saving:

```{r}
#| label: bench-time
boot_mean <- \(i, data) colMeans(data[sample(nrow(data), replace = TRUE), ])

# Without mori — each daemon deserializes a full copy
mirai_map(1:8, boot_mean, .args = list(data = df))[] |> system.time()

# With mori — each daemon maps the same shared memory
mirai_map(1:8, boot_mean, .args = list(data = shared_df))[] |> system.time()

daemons(0)
```

## Installation

``` r
install.packages("mori")
```

## Quick Start

`share()` writes an R object once into shared memory and returns a zero-copy ALTREP view.
Shared objects serialize compactly via ALTREP serialization hooks, working transparently with mirai and any R serialization path.
Shared memory is automatically freed when the object is garbage collected.

```{r}
#| label: quick-start
library(mori)

# Share a vector — returns an ALTREP-backed object
x <- share(rnorm(1e6))
mean(x)

# Serialized form is ~100 bytes, not ~8 MB
length(serialize(x, NULL))
```

## Sharing by Name

`shared_name()` extracts the SHM name from a shared object.
`map_shared()` opens a shared region by name — useful for accessing the same data from another process without serialization:

```{r}
#| label: by-name
x <- share(1:1e6)

# Extract the SHM name
nm <- shared_name(x)
nm

# Another process can map the same region by name
y <- map_shared(nm)
identical(x[], y[])
```

## Use with mirai

Shared objects can be sent to local daemons — the ALTREP serialization hooks ensure only the SHM name crosses the wire, and the daemon maps the same physical memory.

> Workers must run on the same machine — mori shares physical RAM, not bytes over a network.

```{r}
#| label: mirai
library(lobstr)
library(mirai)

daemons(1)

x <- share(rnorm(1e6))

# Worker maps the same shared memory — 0 bytes copied
m <- mirai(list(mean = mean(x), size = lobstr::obj_size(x)), x = x)
m[]

daemons(0)
```

Elements of a shared list also serialize compactly — each element travels as a reference to its position in the parent shared region, not as the full data:

```{r}
#| label: mirai-map
daemons(3)

# Share a list — all 3 vectors in a single shared region
x <- share(list(a = rnorm(1e6), b = rnorm(1e6), c = rnorm(1e6)))

# Each element is sent as (parent_name, index) — zero-copy on the worker
mirai_map(x, \(v) format(lobstr::obj_size(v)))[.flat] |> unique()

daemons(0)
```


## How It Works

### What gets shared

All atomic vector types and lists / data frames are written directly into shared memory, with attributes preserved end-to-end.
Pairlists are coerced to lists.
`share()` returns ALTREP wrappers that point into the shared pages — no deserialization, no per-process memory allocation.

All other R objects (environments, closures, language objects) are returned unchanged by `share()` — no shared memory region is created.

![Diagram showing share() writing an object once into OS-backed shared memory, which is then memory-mapped by other processes using zero-copy ALTREP wrappers](man/figures/mori-diagram.svg)

### Lazy access

A data frame lives in a single shared region; columns are read on demand, so a worker that needs 3 of 100 columns only loads 3.
Character strings are accessed lazily per element.

```{r}
#| label: lazy
df <- share(as.data.frame(matrix(rnorm(1e7), ncol = 100)))
shared_name(df)        # one region for all 100 columns
shared_name(df[[50]])  # sub-path into the same region
```

### Lifetime

Shared memory is managed by R's garbage collector.
The SHM region stays alive as long as any shared object backed by it remains referenced in R — the original returned by `share()`, or a column or sub-list extracted from it, in this or another process.
When no references remain, the garbage collector frees the shared memory automatically.

**Important:** Always assign the result of `share()` to a variable.
The shared memory is kept alive by the R object reference — if the result is used temporarily (not assigned), the garbage collector may free the shared memory before a consumer process has mapped it.

### Copy-on-write

Shared data is mapped read-only.
Mutations are always local — R's copy-on-write mechanism ensures other processes continue reading the original shared data:

- **Structural changes** to a list or data frame (adding, removing, or reordering elements) produce a regular R list.
  The shared region is unaffected.
- **Modifying values** within a shared vector (e.g., `X[1] <- 0`) materializes just that vector into a private copy.
  Other vectors in the same shared region stay zero-copy.

--

Please note that the mori project is released with a [Contributor Code of Conduct](https://shikokuchuo.net/mori/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
